### 2022-03-18 manuscript.tex
Update 01.introduction.md [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/5c8aea3d5fbc147f384068723bceba76aa35d4a9.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/5c8aea3d5fbc147f384068723bceba76aa35d4a9/checks
https://github.com/scherzerthesis/thesis/actions/runs/2004755909

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/d8a34ba19e7004dab8879e50fe9a95405c7d2252/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/d8a34ba19e7004dab8879e50fe9a95405c7d2252}{scherzerthesis/thesis@d8a34ba}
on March 18, 2022.

@@

@@

Due to the strong association between lung adenocarcinoma formation and mutational activation of the MAPK-pathway, many labs have generated genetically-engineered mouse models (GEMMs) that harbor conditionally-activated alleles of either KRAS (G12D) or BRAF(V600E). Expression on BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fail to progress to malignant lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53, and CDKN2A, as loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allow BRAF (V600E) adenomas to progress to adenocarcinoma. These observations are in concordance to the "multiple-hit" hypothesis. This hypothesis states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer". The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels (latest Hanahan and Weinberg). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer, but additional pathways need to be activated to form deadly metastasis.

@@

\textbf{MAPK Signaling}

Normal activation of the MAPK pathways begins with extracellular ligands binding to Receptor-Tyrosine Kinases (RTKs). This leads to the association of the G-protein coupled RAS with GTP, which activates RAS. GTP-loaded RAS recruits and activates RAF kinases (ARAF, BRAF, CRAF) which then phosphorylates and activates MEK kinases which then phosphorylates and activates ERK kinases. This ultimately leads to the activation of a suite of cellular processes that are required for proliferation, differentiation, and cell survival. Therefore, activating mutation in this pathway, from RTKs to kinases, serve to keep the pathway in a constitutively active state. This active state drives many of the hallmarks of cancer that is required for tumor formation. It is important to note that oncogenic activation of the MAPK pathway differs from normal ligand-mediated pathway activation. For example, mitogens, such as EGF, only transiently activate the pathway even with constant ligand stimulation. This immediate and strong spike in pathway activation is integrated by various biological processes in the cell that results in transcription of certain target genes. In contrast, oncogenic signaling thru mutational activation of BRAF(V600E) for example results in constant kinase signaling that is integrated differently than normal pathway stimulation. Thus, an oncogene-specific transcriptional program is activated.

The importance of this pathway is reflected in the strong anti-tumor responses that are seen when small molecule inhibitors are used to target various molecules involved in the MAPK pathway. For example, Dabrafenib and Trametinib, inhibitors of BRAF and MEK kinases, respectfully, are used in patients that harbor BRAF(V600E) melanomas and lung cancers, both offering a clinical benefit (refs). Although there is strong pre-clinical and clinical evidence that BRAF+MEK blockade can lead to tumor regression, not all patients respond to treatment (ref). Therefore, to see a complete response, multiple pathways need to be targeted.

@@

\textbf{Conditional mouse models of BRAF (V600E) lung cancer}

@@

RAS-RAF-MEK-ERK signaling

Talk about the details of raf activation and subsequent regulation of important cellular processes

@@

@@

@@

@@

@@

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/5c8aea3d5fbc147f384068723bceba76aa35d4a9/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/5c8aea3d5fbc147f384068723bceba76aa35d4a9}{scherzerthesis/thesis@5c8aea3}
on March 18, 2022.

@@

\textbf{MAPK signaling in cancer}

standard activation from the top of he MAPK pathway begins with extracellular ligands (EGF, NGF, etc.) binding to Receptor-Tyrosine Kinases (RTKs). This leads to an loading of RAS with GTP, which activates RAS. Many RAS mutations lead to either more efficient loading of GTP or less effciency GTP hydrolysis which both serve to activate RAS. GTP-loaded RAS recruits and activates RAF kinases (ARAF, BRAF, CRAF) which then phosphorylates and activates MEK1/2 kinases which then phosphorylates and activates ERK1/2. This ultimately leads to the activation of a suite of cellular processes that are required for proliferation, differentiation, migration, and cell survival. Therefore, activating mutation in this pathway, from RTKs to kinases, serve to keep the pathway in a constitutively active state. For example, the most common mutation in \emph{BRAF} is V600E. This valine to glutamic acid substitution is located in the kinase domain and mimics phosphorylation, allowing BRAF(V600E) to become independet of upstream regulations like RAS. This active state drives many of the hallmarks of cancer that is required for tumor formation. It is important to note that oncogenic activation of the MAPK pathway differs from normal ligand-mediated pathway activation. For example, mitogens, such as EGF, only transiently activate the pathway even with constant ligand stimulation. This immediate and strong spike in pathway activation is integrated by various biological processes in the cell that results in transcription of certain target genes. In contrast, oncogenic signaling thru mutational activation of BRAF(V600E) results in constant kinase signaling that is integrated differently than normal pathway stimulation. Thus, an oncogene-specific transcriptional program is activated.

The importance of this pathway is reflected in the strong anti-tumor responses that are seen when small molecule inhibitors are used to target various molecules involved in the MAPK pathway. For example, Dabrafenib and Trametinib, inhibitors of BRAF and MEK kinases, respectfully, are used in patients that harbor BRAF(V600E) melanomas and lung cancers(Planchard \emph{et al.}, 2021; Robert \emph{et al.}, 2014). Although there is strong pre-clinical and clinical evidence that BRAF+MEK blockade can lead to tumor regression, not all patients respond to treatment(Tabbò \emph{et al.}, 2021). Therefore, to see a complete response, multiple pathways need to be targeted, such as PI3K or autophagy (Deuker \emph{et al.}, 2014; Truong \emph{et al.}, 2020; Kinsey \emph{et al.}, 2019).

@@

Due to the robust connection between lung adenocarcinomas and mutational activation of genes involved in the MAPK-pathway, such as \emph{EGFR}, \emph{KRAS}, and \emph{BRAF}, many labs have generated genetically-engineered mouse models (GEMMs) that harbor conditionally-activated cancer specific mutant alleles of either EGFR (L858R), KRAS (G12D), or BRAF(V600E). Expression of BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fail to progress to malignant lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53, and CDKN2A, as loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allows BRAF (V600E) adenomas to progress to adenocarcinoma. These observations are in concordance to the "multiple-hit" hypothesis. This hypothesis states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer". The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels (Hanahan, 2022). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer and additional pathways need to be activated to form deadly metastasis.

@@

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-WyV78aHe}{}}%
Deuker,M.M. \emph{et al.} (2014) \href{https://doi.org/10.1158/2159-8290.cd-14-0856}{PI3'-kinase inhibition forestalls the onset of MEK1/2 inhibitor resistance in BRAF-mutated melanoma.} \emph{Cancer Discov}, \textbf{5}, 143--53.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kgaUYMuT}{}}%
Hanahan,D. (2022) \href{https://doi.org/10.1158/2159-8290.cd-21-1059}{Hallmarks of Cancer: New Dimensions.} \emph{Cancer Discov}, \textbf{12}, 31--46.

@@

\leavevmode\vadjust pre{\hypertarget{ref-4S1jzORt}{}}%
Planchard,D. \emph{et al.} (2021) \href{https://doi.org/10.1016/j.jtho.2021.08.011}{Phase 2 Study of Dabrafenib Plus Trametinib in Patients With BRAF V600E-Mutant Metastatic NSCLC: Updated 5-Year Survival Rates and Genomic Analysis.} \emph{J Thorac Oncol}, \textbf{17}, 103--115.

@@

\leavevmode\vadjust pre{\hypertarget{ref-CgOGrXoS}{}}%
Robert,C. \emph{et al.} (2014) \href{https://doi.org/10.1056/nejmoa1412690}{Improved overall survival in melanoma with combined dabrafenib and trametinib.} \emph{N Engl J Med}, \textbf{372}, 30--9.

@@

\leavevmode\vadjust pre{\hypertarget{ref-fXe5uEAl}{}}%
Tabbò,F. \emph{et al.} (2021) \href{https://doi.org/10.1016/j.ctrv.2021.102335}{How far we have come targeting BRAF-mutant non-small cell lung cancer (NSCLC).} \emph{Cancer Treat Rev}, \textbf{103}, 102335.

\leavevmode\vadjust pre{\hypertarget{ref-jKCnCnG4}{}}%
Truong,A. \emph{et al.} (2020) \href{https://doi.org/10.1158/1078-0432.ccr-20-1675}{Chloroquine Sensitizes }. \emph{Clin Cancer Res}, \textbf{26}, 6374--6386.
